---
title: "Stat 652 Project"
author: "Vincent Chiu"
date: '2019-10-16'
output: pdf_document
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading Libraries
```{r}
library(tidyverse)
```


## Loading the data
```{r}
library(nycflights13)
library(Hmisc)
set.seed(42)
original_data <- read_csv("fltrain.csv.gz")
DF <- original_data
```


prof considers ignoring stuff in planes table.

very little precipitation
no way to get info about preceding flight
scheduled landinging time minus departure time might be good


is both sched arrival time and dep time in the same tz?


```{r}
fltrain <- DF
fltrain
```


```{r}

```

prof points: 

very few 

ie common for apeople with a lot of money to not report their income, ie can be biased
if you drop people with lots of income,  not ignorable not response

if there is missing at random; then you can impute without bias


should i impute model?  there is no information? 

cannot measure bias?


potential limitations is introducing bias from not ignorable not response from including model?




can drop the outliers? yes

we have very little chance of predicting magnitude of flight delay.

can we translate and take log scale?


hadley wickham also looked at this data.

replace data to rank?   


fl <- fl %>% mutate(dep_delay=rank(dep_delay))?
going to treat rank as continuous variable.   

replace observations with normal score qqnorm(dep_delay)


#try predicting on rank to see what happens.




